---
title: "Practical Time Series Analysis"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

# Week-1

## Datasets

```{r}
# install.packages("astsa")
require(astsa)
plot(jj, type='o',
main='Johnson&Johnson quaterly earnings per share',
xlab='Quarters', ylab='Earnings')
```

```{r}
plot(astsa:: flu,
main='Monthly pneumonia and influenza deaths in the U.S.',
xlab='Months', ylab='Deaths per 10,000 people')
```

```{r}
plot(astsa:: globtemp, type='o',
main='Global mean land-ocean deviations average temperature of 1951-1980',
xlab='Years', ylab='Temperature deviations')
```

```{r}
plot(astsa:: globtempl, type='o',
main='Global mean land deviations average temperature of 1951-1980',
xlab='Years', ylab='Temperature deviations')
```

```{r}
plot(astsa:: star,
main='The magnitude of a star taken at midnight for 600 consecutive days',
xlab='Days', ylab='Magnitude')
```

## Notation

-   $\mathrm{E}$: expected value

-   $\mu$: average (expected estimation)

-   $\mathrm{V}$, $\mathrm{Var}$ : variance

-   $\sigma$: standard error (estimated root squared variance)

-   $\rho$: correlation

-   $r$: correlation estimation

-   $\mathrm{cov}$: covariance

-   $k$, $\tau$: lag

-   $IID$, $iid$: independent, identically distributed

-   $\mathcal{N}$: normal distribution

-   $\epsilon$: error

-   $\mathrm{\gamma}$: autocovariance function

-   $X_t$: discrete

-   $X(t)$: continuous

## (weak) Stationarity

1.  no systematic changes in the mean (no trend, not even piecewise)
2.  no systematic changes in variance
3.  no periodic fluctuations (seasonality)

$$
\mathrm{E}[\epsilon_i] = 0 \\
\mathrm{V}[\epsilon_i] = \sigma^2{\epsilon_i} = \text{constant} = \sigma^2 \\
\mathrm{cov}[\epsilon_i,\epsilon_j] = \sigma\{\epsilon_i, \epsilon_j\} = 0 \quad \forall i \neq j
$$

## Regression Model assumptions

When we find estimates of the slope and intercept using, for example
Ordinary Least Squares (OLS), we are not really making any
distributional assumptions, just taking a cloud of points and finding
numbers that we call a slope and an intercept that minimize a quality
term.

If we wish to perform some inferences (confidence intervals, hypothesis
tests), then we need to make distributional assumptions. Generally, we
make the (often reasonable!) assumptions that the errors are (weak)
stationary) and normal distributed, e.g., the errors

1.  are normally distributed and, on average, zero;
2.  all have the same variance (they are homoscedastic), and
3.  are unrelated to each other (they are independent across
    observations).

$$
\epsilon_i \stackrel{iid}{\sim} \mathcal{N} ( \mu=0, \; \sigma^2 = \text{constant} )
$$

## Variance

It measures the "dispersion" for 1 random variable

$$
\sigma^2 \equiv \mathrm{V}[X] \equiv \mathrm{E}[ (X - \mu_X) (X - \mu_X) ]
$$

$$
V[X+Y] = V[X] + V[Y] + 2\;\mathrm{cov}(X,Y) \\
V[aX] = a^2 V[X]
$$

## Covariance

A linear dependence measure between 2 random variables

$$
\mathrm{COV}[X, Y] \equiv \mathrm{E}\left[ (X - \mu_X) (Y - \mu_Y) \right] \\
\mathrm{COV}[X, Y] = \mathrm{E}\left[ X \cdot Y \right] - \mathrm{E}\left[ X  \right] \mathrm{E}\left[  Y \right]$$

Autocovariance

$$
\mathrm{COV}[X, X] = \mathrm{E}\left[ (X - \mu_X) (X - \mu_X) \right] = V[X] = \sigma_X^2
$$

## Correlation

Like an adimensional covariance

$$
\mathrm{\rho}[X, Y] \equiv \mathrm{E}\left[ \left(\frac{X - \mu_X}{\sigma_X}\right) \left(\frac{Y - \mu_Y}{\sigma_Y}\right) \right]
$$

## Stochastic process

A collection of random variables that follows a distribution (let's say
a normal one):

$$
X_t \sim \mathcal{N} ( \mu, \; \sigma^2 ) 
$$

An individual trajectory is a realization of a stochastic process, our
time series.

The set of all possible trajectories is called the ensemble.

## Mean function

$$
\mu(t) \equiv \mu_t \equiv E[X(t)]
$$

## Variance function

$$
\sigma^2(t) \equiv \sigma^2_t \equiv V[X(t)]
$$

## Autocovariance function

$$
\gamma(s,t) = \mathrm{Cov}[X_s,X_t] = \mathrm{E}\left[ (X_s - \mu_s) (X_t - \mu_t) \right] 
$$

In particular

$$
\gamma(t,t) = \mathrm{E}\left[ (X_t - \mu_t)^2 \right] = \mathrm{V}[X_t] = \sigma_t^2
$$

For a process with lag k

$$
\mathrm{COV}[X_t, X_{t+k}] = \mathrm{E}\left[ X_{t} \cdot X_{t+k} \right] - \mathrm{E}\left[ X_{t} \right] \mathrm{E}\left[  X_{t+k} \right]
$$

## Correlation

## Autocorrelation function (ACF)

variable and lagged variable

-   weak stationarity

AC coefficient $$ -1 \le \rho_k - \frac{\gamma_k}{\gamma_0} \le 1 $$

Estimation of autocorrelation coefficient $$ r_k = \frac{c_k}{c_0}  $$

Any time series has correlation 1 with itself, i.e., autocorrelation at
lag 0 is 1.

## Weak stationary discrete process

None assumptions about distributions.

1.  Mean function is constant
2.  Autocovariance function depends only from the lag $\tau$
3.  From 2, the variance is also constant

$$
\mu(t) = \mu  \\
\gamma(t_1,t_2) = \gamma(t_2-t_1) = \gamma(\tau) \\
\implies \gamma(t,t) = \gamma(\tau=0) = \sigma^2(t) = \sigma^2
$$

Let's linearly combine the current and a lagged value and compute its
variance (since the latter is a squared term, it is therefore
non-negative)

$$
V[a X (t) + b X (t+\tau)] \geq 0$$

After we apply variance properties

$$
a^2 V[ X (t) ] + b^2 V[ X (t+\tau) ] + 2 ab \;\mathrm{cov}\left( X(t),X(t+\tau) \right) \geq 0
$$

For a weak stationary process, the variance is constant

$$
a^2 \sigma^2 + b^2 \sigma^2 + 2 ab \;\mathrm{cov}\left( X(t),X(t+\tau) \right) \geq 0 
$$

For $a=b=1$:

$$
\sigma^2 + \sigma^2 + 2 \;\mathrm{cov}\left( X(t),X(t+\tau) \right)  \geq 0 \\ \sigma^2 +  \mathrm{cov}\left( X(t),X(t+\tau) \right) \geq 0 \\
\gamma(0) + \gamma(\tau) \geq 0 \\
\frac{\gamma(0)}{\gamma(0)} + \frac{\gamma(\tau)}{\gamma(0)} \geq 0 \\
1 + \rho(\tau) \geq 0 \\
\rho(\tau) \geq -1
$$

For $a=1,\;b=-1$:

$$
\sigma^2 + \sigma^2 - 2 \;\mathrm{cov}\left( X(t),X(t+\tau) \right)  \geq 0 \\ \sigma^2 -  \mathrm{cov}\left( X(t),X(t+\tau) \right) \geq 0 \\
\gamma(0) - \gamma(\tau) \geq 0 \\
\frac{\gamma(0)}{\gamma(0)} - \frac{\gamma(\tau)}{\gamma(0)} \geq 0 \\
1 - \rho(\tau) \geq 0 \\
1 \geq \rho(\tau)
$$

Therefore

$$ -1 \le \rho(t) \le 1 $$

## Strict stationarity

The joint distribution of the full set of random variables is usually
unknown.

Joint distribution of $X(t_1), \cdots, X(t_k)$ is the same as the joint
distribution of $X(t_1+\tau), \cdots, X(t_k+\tau)$

Implications

-   \$\mathcal{N} (X(t_1)) \equiv \\mathcal{N} (X(t_1+\tau)) \$

-   random variables identically distributed (but not necessarily
    independent)

-   $\mu(t)=\mu$

-   $\sigma^2(t) = \sigma^2$

-   $\gamma(t_1,t_2) = \gamma(t_2-t_1) = \gamma(\tau)$

## White noise

$$ 
X_t \stackrel{iid}{\sim} \mathcal{N} ( \mu=0, \; \sigma^2 = \text{constant} )
 $$

Mean function: \$ \mu(t) = \mu = 0 \$

Autocovariance function:

\#TODO: ENTENDER AUTOCOVARIANCE

$$
    \gamma(t_1,t_2)= 
\begin{cases}
    0           & t_1 \neq t_2 \\
    \sigma^2    & t_1 = t_2
\end{cases}
$$

Autocorrelation function:

$$
    \rho(t_1,t_2)= 
\begin{cases}
    0    & t_1 \neq t_2 \\
    1    & t_1 = t_2
\end{cases}
$$

As the mean and variance are constant along the time and there is no
seasonality, it is a weak stationary process.

## Random walk

$$ X_t = X_{t-1} + Z_t(\mathcal{N}(\mu=0,\sigma^2=1)) $$

Random walk is the accumulation of random deviations from previous steps
until the current time.

$$ 
X_1 = Z_1 \\
X_2 = X_1 + Z_2 = Z_1 + Z_2 \\
X_3 = X_2 + Z_3 = Z_1 + Z_2 + Z_3
$$

For random variables, the mean of the sum is the sum of the means:

$$
\mathrm{E}[X_t] = \mathrm{E}\left[ \sum_{i=1}^{t} Z_i \right] = \sum_{i=1}^{t} \mathrm{E}\left[  Z_i \right] = t \mu
$$

For independent random variables, the variance of the sum is the sum of
the variances.

$$
\mathrm{V}[X_t] = \mathrm{V}\left[ \sum_{i=1}^{t} Z_i \right] = \sum_{i=1}^{t} \mathrm{V}\left[  Z_i \right] = t \sigma^2
$$

So random walk is not a stationary process, since even when $\mu=0$, its
variance increase over time

One can obtain a stationary stochastic process from the random walk
using the difference operator.

## Moving average

The current value of the process now is a linear combination of the
noises from current and past time steps. Autocorrelation function of the
process cuts off and becomes zero at the order $q$ of the process.

$$ 
Z_t \stackrel{iid}{\sim} \mathcal{N} ( E[Z_t]=0, \; V[Z_t]=\sigma_Z^2 )
 $$

$$ MA(q)\text{ process:}\qquad X_t = \sum_{i=0}^q {\beta_i Z_{t-i}} $$

Mean function

$$
E[X_t] = \beta_0 E[Z_t] + \beta_1 E[Z_{t-1}] +\cdots+ \beta_q E[Z_{t-q}] = 0
$$

Variance function

$$
V[X_t] = \beta_0^2 V[Z_t] + \beta_1^2 V[Z_{t-1}] +\cdots+ \beta_q^2 V[Z_{t-q}] = \sigma_Z^2 \sum_{i=0}^q {\beta_i^2}
$$

Covariance function, lag k

$$
\mathrm{cov}[X_t, X_{t+k}] = 
\mathrm{cov}\left[ 
\beta_0 Z_t + \beta_1 Z_{t-1} +\cdots+ \beta_q Z_{t-q}, \\
\qquad \qquad \qquad \qquad \qquad \quad \beta_0 Z_{t+k} + \beta_1 Z_{t+k-1} +\cdots+ \beta_q Z_{t+k-q} \right]
$$

$$
\mathrm{COV}[X_t, X_{t+k}] = \mathrm{E}\left[ X_{t} \cdot X_{t+k} \right] - \mathrm{E}\left[ X_{t} \right] \mathrm{E}\left[  X_{t+k} \right]
$$

As the mean function is zero.

$$
\mathrm{E}\left[ X_{t} \right] =  \mathrm{E}\left[  X_{t+k} \right] = 0 \implies
\mathrm{COV}[X_t, X_{t+k}] = \mathrm{E}\left[ X_{t} \cdot X_{t+k} \right]
$$

$$
E\left[ \left( 
\beta_0 Z_t + \beta_1 Z_{t-1} +\cdots+ \beta_q Z_{t-q} \right) \cdot
\left( \beta_0 Z_{t+k} + \beta_1 Z_{t+k-1} +\cdots+ \beta_q Z_{t+k-q} \right)  \right] \\
= E\begin{bmatrix}
\beta_0\beta_0 Z_{t}Z_{t+k} + \beta_0 \beta_1 Z_{t}Z_{t+k-1} +\cdots+ \beta_0\beta_q Z_{t}Z_{t+k-q} \\
+\\
\beta_1\beta_0 Z_{t-1}Z_{t+k} + \beta_1 \beta_1 Z_{t-1}Z_{t+k-1} +\cdots+ \beta_1\beta_q Z_{t-1}Z_{t+k-q} \\
+\\
\beta_q\beta_0 Z_{t-q}Z_{t+k} + \beta_q \beta_1 Z_{t-q}Z_{t+k-1} +\cdots+ \beta_q\beta_q Z_{t-q}Z_{t+k-q}
\end{bmatrix}
$$

$$
    \mathrm{E}\left[ Z_{t} \cdot Z_{t+i} \right] =
\begin{cases}
    \mathrm{E}\left[ Z_{t} \right] \cdot \mathrm{E}\left[ Z_{t+i} \right] = 0           & t \neq t+i \\
    \mathrm{E}\left[ Z_{t} \right] \cdot \mathrm{E}\left[ Z_{t} \right] = V[Z_{t}] = \sigma^2    & t = t+i
\end{cases}
%= \mathrm{E}\left[ X_{t} \right]  \cdot \mathrm{E}\left[ X_{t+k} \right] =
$$

For $k=0$, only at the main diagonal we refer to the same instant.

$$
\mathrm{COV}[X_t, X_{t}] = E\begin{bmatrix}
\beta_0\beta_0 Z_{t}Z_{t} + 0 +\cdots+ 0 \\
+\\
0 + \beta_1 \beta_1 Z_{t-1}Z_{t-1} +\cdots+ 0 \\
+\\
0 + 0 +\cdots+ \beta_q\beta_q Z_{t-q}Z_{t-q}
\end{bmatrix}
= \sigma^2 \sum_{i=0}^{q} \beta_i^2
$$

For $k=1$, only a shift diagonal contains terms at the same instant.

$$
\mathrm{COV}[X_t, X_{t}] = E\begin{bmatrix}
0 + \beta_0\beta_1 Z_{t}Z_{t} +\cdots+ 0 \\
+\\
0 + 0 + \beta_1 \beta_2 Z_{t-1}Z_{t-1} +\cdots+ 0 \\
+\\
0 + 0 +\cdots+ 0
\end{bmatrix}
= \sigma^2 \sum_{i=0}^{q-1} \beta_i\beta_{i+1}
$$

For $k=q$,

$$
\mathrm{COV}[X_t, X_{t}] = E\begin{bmatrix}
0 + 0 +\cdots+ \beta_0\beta_q Z_{t}Z_{t+q-q} \\
+\\
0 + 0  +\cdots+ 0 \\
+\\
0 + 0 +\cdots+ 0
\end{bmatrix}
= \sigma^2 \beta_0\beta_q
$$

Generic $k\leq q$, inspecting the coefficients

$$
\mathrm{cov} [X_t, X_{t+k}] = 
\mathrm{cov} \left[ \left( 
\beta_0 Z_t + \beta_1 Z_{t-1} +\cdots+ \beta_q Z_{t-q} \right) \cdot
\left( \beta_0 Z_{t+k} + \beta_1 Z_{t+k-1} +\cdots+ \beta_q Z_{t+k-q} \right)  \right]
$$

$$
\begin{matrix}
Z_{t-q} & Z_{t-(q-1)} & \cdots & Z_{t-(q-k)} & \cdots & Z_{t-1} & Z_{t} & \cdots & & \\
\beta_{q} & \beta_{q-1} & \cdots & \beta_{q-k} & \cdots & \beta_{1} & \beta_{0} & \cdots & 0 & 0 \\
 &  & \cdots & Z_{t+k-q} & \cdots & Z_{t+k-(k+1)} & Z_{t+k-k} & \cdots & Z_{t+k-1} & Z_{t+k} \\
0 & 0 & \cdots & \beta_{q} & \cdots & \beta_{k+1} & \beta_{k} & \cdots & \beta_{1} & \beta_{0} \\
\end{matrix}
$$

$$
\mathrm{cov} [X_t, X_{t+k}] = \sigma^2 \sum_{i=0}^{q-k}{\beta_i \beta_{i+k}}
$$

$$
    \gamma(t_1,t_2) = \gamma(k) =
\begin{cases}
    0           & k > q \\
    \sigma_Z^2 \sum_{i=0}^{q-k}{\beta_i \beta_{i+k}}    & k \leq q
\end{cases}
$$

Since the mean function is constant and the autocovariance function does
not depend on time (it depends only from the lag spacing), the moving
average is a weak stationary process.

When there is no lag, $k=0$:

$$
\gamma(k=0) = \sigma_Z^2 \sum_{i=0}^{q}{\beta_i \beta_{i}} 
$$

Therefore the autocorrelation function is

$$
\rho(k) = \frac{\gamma(k)}{\gamma(0)} 
= \frac{\sigma_Z^2 \sum_{i=0}^{q-k}{\beta_i \beta_{i+k}}}{\sigma_Z^2 \sum_{i=0}^{q}{\beta_i \beta_{i}} }
= \frac{\sum_{i=0}^{q-k}{\beta_i \beta_{i+k}}}{\sum_{i=0}^{q}{\beta_i^2}}
$$

Theoretically, it is 0 starting at lag 4. But for a time series, it will
be some value which is non-significant. Blue lines are significance
level

```{r}
set.seed(2^10)
z=NULL
z=rnorm(1000)
data=NULL
for(i in 4:1000){
  data[i-3]=z[i]+0.2*z[i-1]+0.3*z[i-2]+0.4*z[i-3]
}
data=ts(data)
# find acf below
# blue lines are significance level
acf(data)
```

```{r}
par( mfrow=c(3,1) );
plot( arima.sim(n=150, list(order=c(0,0,0) )  ), main="WN" );
plot( arima.sim(n=150, list(ma=c(0.33, 0.33, 0.33)      )  ) , main="MA3");
plot( arima.sim(n=150, list(ma=c(0.2, 0.2, 0.2, 0.2, 0.2) )  ), main="MA5" );
```

I don't get the 0.531 as $\rho(1)$

```{r}
set.seed=1
(acf(arima.sim(n=1000, model=list(ma=c(0.5,0.5)))))
```

## Summary

+---------------+------------+------------+------------+------------+
| Function      | Mean       | Aut o      | Variance   | Stationary |
|               |            | covariance |            |            |
| Process       |            |            |            |            |
+===============+============+============+============+============+
| **Strict s    | constant   | lag        | constant   | Strict     |
| tationary**   |            | dependent  |            |            |
+---------------+------------+------------+------------+------------+
| **Weak s      | constant   | depends    | constant   | Weak       |
| tationary**   |            | only from  |            |            |
|               |            | lag        |            |            |
+---------------+------------+------------+------------+------------+
| **Random      | p r        | p r        | **p r      | No         |
| walk**        | oportional | oportional | oportional |            |
|               | to t       | to t       | to t**     |            |
+---------------+------------+------------+------------+------------+
| **White       | constant   | constant   | constant   | Strict /   |
| noise**       |            |            |            | weak       |
+---------------+------------+------------+------------+------------+
| **Moving      |            |            |            | weak       |
| average**     |            |            |            |            |
+---------------+------------+------------+------------+------------+

# Week-2

## Autoregressive process

Order p

$$
X_t=Z_t+history
$$

White noise, innovations, random disturbance

$$
Z_t\stackrel{iid}{\sim}(\mathcal{N}(\mu=0,\sigma^2=1))
$$

History

$$
history = \phi_1 X_{t-1} + \cdots + \phi_p X_{t-p}
$$

$$
AR(p): X_t = Z_t + \phi_1 X_{t-1} + \cdots + \phi_p X_{t-p} \\
MA(q): X_t = \theta_0 Z_{t} + \theta_1 Z_{t-1} + \cdots + \theta_q Z_{t-q}
$$

MA: finite set of innovations $Z_t$'s

AR: current innovation $Z_t$ and history of a finite set of prior states
$X_t$'s

### Random walk

$$
X_t=X_{t-1}+Z_t
$$

For $p=\phi_1=1$

$$
X_t=X_{t-1}+Z_t
$$

```{r}
set.seed(2016); N=1000; phi = 1;
Z = rnorm(N,0,1); X=NULL; X[1] = Z[1];
for (t in 2:N) {
X[t] = Z[t] + phi*X[t-1] ;
}
X.ts = ts(X)
par(mfrow=c(2,1))
plot(X.ts,main="AR(1) Time Series on White Noise, phi=1")
X.acf = acf(X.ts, main="AR(1) Time Series on White Noise, phi=1")
```

First order AR(1)

```{r}
set.seed(2016); N=1000; phi = .4;
Z = rnorm(N,0,1); X=NULL; X[1] = Z[1];
for (t in 2:N) {
X[t] = Z[t] + phi*X[t-1] ;
}
X.ts = ts(X)
par(mfrow=c(2,1))
plot(X.ts,main="AR(1) Time Series on White Noise, phi=.4")
X.acf = acf(X.ts, main="AR(1) Time Series on White Noise, phi=.4")

(r.coef = X.acf$acf)
(c (mean(X), var(X)))
```

```{r}
for (phi1 in c(0.1, 0.9)) {
  X.ts <- arima.sim(list(ar = c(phi1)), n=1000)
  par(mfrow=c(2,1))
  plot(X.ts,main=paste("AR(1) Time Series, phi1=",phi1))
  X.acf = acf(X.ts, main="Autocorrelation of AR(1) Time Series")
}
```

Second order AR(2)

Coefficients range that assure stationarity

$$
-1 < \phi_2 <1 \\
\phi_2 < 1 + \phi_1 \\
\phi_2 < 1 - \phi_1
$$

```{r}
X.ts <- arima.sim(list(ar = c(.7, .2)), n=1000)
par(mfrow=c(2,1))
plot(X.ts,main="AR(2) Time Series, phi1=.7, phi2=.2")
X.acf = acf(X.ts, main="Autocorrelation of AR(2) Time Series")
```

```{r}

phi1 = .4; phi2 = .3;

par(mfrow=c(2,1))
X.ts <- arima.sim(list(ar = c(phi1, phi2)), n=1000)
plot(X.ts,main=paste("AR(2) Time Series, phi1=",phi1,"phi2=",phi2))
acf(X.ts,main="ACF")

```

## Backshift operator

$$
AR(p): X_t = Z_t + \phi_1 X_{t-1} + \cdots + \phi_p X_{t-p}
$$

$$
X_{t-1} = B^1 X_{t} \\
X_{t-2} = B^2 X_{t} \\
\vdots \\
X_{t-p} = B^p X_{t}
$$

$$
X_t = Z_t + \phi*1 B X{t} + \cdots + \phi*p B^p \\
X{t} = Z_t + (\phi_1 B + \cdots + \phi*p B^p) X{t} \\
Z_t = (1 - \phi_1 B - \cdots - \phi*p B^p) X{t} = \Phi(B) X_t \\
X_t = \Phi^{-1}(B) Z_t = \frac{1}{\Phi(B)}Z_t \\
X_t = \frac{1}{(1 - \phi_1 B  + \cdots + \phi_p B^p)}Z_t = \frac{1}{1 - ( \phi_1 B  + \cdots + \phi_p B^p )} Z_t
$$

$$
MA(q\to\infty): X_t = \theta_0 Z_{t} + \theta_1 Z_{t-1} + \cdots + \theta_q Z_{t-q} \\
 X_t = ( \theta_0 +  \theta_1 B + \theta_2 B^2 + \cdots ) Z_{t}
$$

MA(q) with a drift

$$
X_t = \mu +  \theta_0 Z_{t} + \theta_1 Z_{t-1} + \cdots + \theta_q Z_{t-q} \\
X_t - \mu = \Theta(B) Z_t \\
\Theta(B) = \theta_0 + \theta_1 B + \cdots +  \theta_q B^q
$$

Ex:

MA(2):

$$
X_t = Z_t + 0.2 Z_{t-1} + 0.04 Z_{t-2} = (1+0.2B+0.04B^2)Z_t = \beta(B)Z_t
$$

First order

$$
AR(1): X_t = Z_t + \phi B X_{t} = Z_t + \phi X_{t-1} \\
X_t = Z_t + \phi ( Z_{t-1} + \phi X_{t-2} = Z_t + \phi X_{t-1} + \phi^2 X_{t-2} \\
( 1 - \phi B ) X_{t} = Z_t \\
X_t = \frac{1}{( 1 -  \phi B )}Z_t = ( 1 + \phi B + \phi^2 B^2 + \cdots ) Z_t \\
X_t = Z_t + \phi Z_{t-1} + \phi^2 Z_{t-2} + \cdots
$$

AR(p), $E[Z_t] = 0, V[Z_t]=\sigma_Z^2$, \$V[aX] = a\^2 V[X] \$

$$
E[X_t] = E [ ( 1+\theta_1 B + \theta_2 B^2 + \cdots  ) Z_t ] = E [ Z_t ] + \theta_1 E [ Z_{t-1} ] + \cdots + \theta_k E [ Z_{t-k}] + \cdots   = 0
$$

$$
V[X_t] = V [ ( 1+\theta_1 B + \theta_2 B^2 + \cdots  ) Z_t ] = V [ Z_t ] + \theta_1^2 V [ Z_{t-1} ] + \cdots + \theta_k^2 V [ Z_{t-k}] + \cdots   = \sigma_Z^2(1+\theta_1^2+\cdots+\theta_k^2+\cdots) = \sigma_Z^2 \sum_{i=0}^{\infty}{\theta_i^2}
$$

Ex:

AR(2):

$$
X_t = Z_t + 0.2 X_{t-1} + 0.3 X_{t-2} = Z_t + (0.2B+0.3B^2)X_t \\
Z_t = (1-0.2B-0.3B^2)X_t = \phi(B)X_t
$$

Necessary condition for stationarity. infinite series to converge.

$$
\sum_{i=0}^{\infty}{ | \theta_i | }
$$

$$
MA(q): \gamma(k) = \sigma_Z^2 \sum_{i=0}^{q-k}{\theta_i \theta_{i+k}} \\
AR(p): \gamma(k) = \sigma_Z^2 \sum_{i=0}^{\infty}{\theta_i \theta_{i+k}}
$$

$$
\rho (k) = \phi_1^k
$$

$$
\text{White noise}: \phi = 0 \\
\text{Random walk}: \phi = 1 \\
\text{Alternating signal correlation}: \phi < 1\\
\text{First order stationary}: -1 < \phi < 1
$$

## Series

Sequence of ordered numbers.

Convergent if $\lim_{n\to\infty}a_n=a$

Ex:

$$
a_n = \frac{n}{n+1}:\quad \frac{1}{1+1}=\frac{1}{2}, \frac{2}{2+1}=\frac{2}{3}, \cdots \to 1 \\
a_n = 3^n:\quad 3^1=3, 3^2=9, \cdots \to \infty \\
a_n = \sqrt{n}:\quad \sqrt{1}=1, \sqrt{2}, \cdots \to \infty \\
a_n = \frac{1}{n^2}:\quad \frac{1}{1^2}=1, \frac{1}{2^2}=\frac{1}{4}, \cdots \to 0
$$

Partial sums and convergent series sum

$$
s_n = \sum_{i=1}^{n} a_i, \quad s = \sum_{i=1}^{\infty} a_i
$$

Convergent partial sums \$\implies\$ convergent infinite series.

$$
\sum_{i=1}^{\infty} \frac{1}{2^i} = 1 \\
\sum_{i=1}^{\infty} \frac{1}{i^2} = \frac{\pi^2}{6} \\
\sum_{i=1}^{\infty} \frac{(-1)^{i+1}}{i} = \ln(2)
$$

Divergent

$$
\sum_{i=1}^{\infty} 3^i \\
\sum_{i=1}^{\infty} (2i+1) \\
\sum_{i=1}^{\infty} \frac{1}{i}
$$

$$
a_n = \frac{a_1}{1-r} =   \sum_{i=1}^{\infty} a_1 r^{i-1}
$$

Ex:

$$
\sum_{i=1}^{\infty} \frac{1}{2^i} = \frac{1}{2^1} + \frac{1}{2^2} + \cdots
$$

$$
a_n = \frac{1}{2^i} \\
a_1 = \frac{1}{2^1} = \frac{1}{2} \\
r = \frac{a_2}{a_1} = \frac{1}{2^2}\frac{2^1}{1} = \frac{1}{2} \\
a_1 r^{i-1} = \frac{1}{2} \frac{1}{2^{i-1}} = \frac{1}{2^{i}} \\
\sum_{i=1}^{\infty} a_1 r^{i-1} = \frac{a_1}{1-r} = \frac{\frac{1}{2}}{1-\frac{1}{2}} = \frac{\frac{1}{2}}{\frac{1}{2}} = 1
$$

Ex:

Absolute convergence

$$
\sum_{k=1}^{\infty} | a_k |
$$

## Geometric series

$$
\frac{1}{1-a} = a^0 + a^1 + a^2 + \cdots = \sum_{i=0}^{\infty}a^i, \qquad |a| < 1
$$

$$
|a|: 
\begin{cases}
<1 & \text{convergent}  & \text{Ex: } a=\frac{1}{2}; & 1 + \frac{1}{2} + \frac{1}{4} + \frac{1}{8} \cdots \to \frac{1}{1-\frac{1}{2}} = 2 \\
\neq 0 & \text{zero denominator} \\
>1 & \text{divergent}  & \text{Ex: } a=2; & 1 + 2 + 4 + 8 + \cdots \to \infty \end{cases}
$$

## Invertibility

-   $X_t$: stochastic process

-   $Z_t$: innovations (random disturbances, white noise)

-   $X_t$: invertible, if $Z_t=\sum_{k=0}^{\infty}\pi_{k}X_{t-k}$, where
    $\sum_{k=0}^{\infty}|\pi_{k}|$ convergent.

Model 1

$$
X_t = Z_t + 2 Z_{t-1} \\
\gamma(k) = cov[X_{t+k},X_t] = cov[Z_{t+k} + 2 Z_{t+k-1},Z_t + 2 Z_{t-1}]
$$

$$
k: 
\begin{cases}
<0 & \gamma(k<0) = \gamma(-k>0)  \\
= 0 & \gamma(k=0) = cov[Z_{t} + 2 Z_{t-1},Z_t + 2 Z_{t-1}] = cov[Z_{t}, Z_t] + 2^2 cov[Z_{t-1}, Z_{t-1}] = \sigma_Z^2 + 4 \sigma_Z^2 = 5 \sigma_Z^2 \\
=1 & \gamma(k=1) = cov[Z_{t+1} + 2 Z_{t}, Z_t + 2 Z_{t-1}] = cov[2 Z_{t}, Z_t] = 2  \sigma_Z^2 \\
>1 & \gamma(k>1) = 0, \quad \text{since } t+k > t+k-1 > t > t-1 \implies \text{uncorrelated } Z\text{'s}
\end{cases}
$$

$$
\gamma(k) =
\begin{cases}
\gamma(-k) & k<0  \\
5 \sigma_Z^2 & k=0  \\
2 \sigma_Z^2 & k=1  \\
0 & k>1
\end{cases}
$$

$$
\rho(k)=\frac{\gamma(k)}{\gamma(0)} = \begin{cases}
\rho(-k) & k<0  \\
1 & k=0  \\
\frac{2}{5} & k=1  \\
0 & k>1
\end{cases} 
$$

Divergent

$$
\sum_{k=0}^{\infty}|\pi_{k}| = \sum_{k=0}^{\infty} 2^{k}
$$

```{r}
theta = c(1.0,2.0)
par( mfrow=c(2,1) );
plot( arima.sim(n=150, list(ma=theta)  ) , main="MA2");

set.seed=1
(acf(arima.sim(n=1000, model=list(ma=theta))))
```

Model 2

$$
X_t = Z_t + \frac{1}{2} Z_{t-1} \\
\gamma(k) = cov[X_{t+k}, X_t] 
= cov[ Z_{t+k} + \frac{1}{2} Z_{t+k-1}, Z_t + \frac{1}{2} Z_{t-1}] \\
\gamma(0) = cov[ Z_{t} + \frac{1}{2} Z_{t-1}, Z_t + \frac{1}{2} Z_{t-1}]
 = cov[Z_{t}, Z_{t}]  + \frac{1}{2} cov[Z_{t-1}, Z_{t-1}] = \sigma_Z^2 + \left(\frac{1}{2}\right)^2\sigma_Z^2 = \sigma_Z^2\left(1+\frac{1}{4}\right) \\
\gamma(1) = cov[ Z_{t+1} + \frac{1}{2} Z_{t}, Z_t + \frac{1}{2} Z_{t-1}]
 = \frac{1}{2} cov[Z_{t}, Z_{t}] = \frac{1}{2}\sigma_Z^2 \\
\rho(1) = \frac{\gamma(1)}{\gamma(0)} = \frac{\frac{1}{2}\sigma_Z^2}{\sigma_Z^2\left(1+\frac{1}{4}\right)} = \frac{1}{2}\frac{4}{5} = \frac{2}{5}
$$

Convergent

$$
\sum_{k=0}^{\infty}|\pi_{k}| = \sum_{k=0}^{\infty}\frac{1}{2^{k}}
$$

```{r}
theta = c(1.0,0.5)
par( mfrow=c(2,1) );
plot( arima.sim(n=150, list(ma=theta)  ) , main="MA2");

set.seed=1
(acf(arima.sim(n=1000, model=list(ma=theta))))
```

MA(1) -\> AR($\infty$)

$$
X_t = Z_t + \beta Z_{t-1} \\
Z_t = X_t - \beta Z_{t-1} = X_t - \beta (X_{t-1} - \beta Z_{t-2} ) = X_t -\beta X_{t-1} +\beta^2 Z_{t-2} \\
Z_t = \sum_{k=0}^{\infty}(-\beta)^kX_{t-k} \\
X_t = Z_t + \sum_{k=1}^{\infty}(-\beta)^{k+1}X_{t-k} \\
$$

Convergence: $|\beta|<1$

$$
X_t = Z_t + \beta Z_{t-1} = Z_t + \beta B Z_{t} = (1+\beta B)Z_{t} = \beta(B)Z_{t} \\
Z_t = \beta(B)^{-1} X_t \\
\beta(B)^{-1} = \frac{1}{(1+\beta B)} = 1 - \beta B + \beta^2 B^2 +\cdots  = \sum_{k=1}^{\infty}(-\beta B)^{k-1}  \\
$$

1\.

$$
-3+\frac{3}{2}-\frac{3}{4}+\cdots 
= -3 \sum_{k=0}^{\infty}\left(-\frac{1}{2}\right)^{k}\\
a \sum_{k=0}^{\infty} r^{k}
= \frac{a}{1-r}, \text{ if } |r|<1, \text{ convergent} \\
S = \frac{-3}{1-\left(-\frac{1}{2}\right)}
= \frac{-3}{\frac{3}{2}}
= -2
$$

2\.

$$
\frac{4}{1+x} 
= \frac{4}{1-(-x)} 
= 4 \sum_{k=0}^{\infty} (-x)^{k}
= 4 - 4x + 4x^2 - \cdots   \\ 
\frac{a}{1-r} = a \sum_{k=0}^{\infty} r^{k}
$$

3\.

$$
X_t = 0.5 X_{t-1} + Z_t + 0.7 Z_{t-1} \\
(1-0.5B)X_t = (1 + 0.7B) Z_{t}
$$

4.  AR(2)

$$
X_t = X_{t-1} + 2 X_{t-2} + Z_t \\
(1-B-2B^2)X_t = Z_{t}
$$

5\.

$$
X_{t}=Z_{t}+3Z_{t-1} = (1+3B)Z_{t} \\
1+3B=0 \implies r=-\frac{1}{3} \\
|r| = | -\frac{1}{3} | = \frac{1}{3} < 1 \implies \text{not invertible}
$$

6\.

$$
X_t = Z_t - \theta Z_{t-1} - 6 \theta^2 Z_{t-2} = (1-\theta B-6\theta^2B^2) Z_t \\
1-\theta B-6\theta^2B^2 = 0 \\
S = -\frac{b}{a} = -\frac{-\theta}{-6\theta^2} = -\frac{1}{6\theta} \\
P = \frac{c}{a} = \frac{1}{-6\theta^2} = -\frac{1}{6\theta^2} \\
r_1 = -\frac{1}{2\theta} = -(2\theta)^{-1},\quad r_2 = \frac{1}{3\theta} = (3\theta)^{-1} \\
|r| > 1  \implies \text{invertible} \\
|r_1| = |-(2\theta)^{-1}| = |2\theta|^{-1} > 1 \implies |\theta| < \frac{1}{2} \\
|r_2| = |(3\theta)^{-1}| = |3\theta|^{-1}  > 1 \implies |\theta| < \frac{1}{3} \\
\implies |\theta| < \frac{1}{3}
$$

7.AR(2)

$$
X_t = X_{t-1} + 2 X_{t-2} + Z_t \\
(1-B-2B^2)X_t = Z_{t} \\
1-B-2B^2 = 0 \\
S = -\frac{b}{a} = -\frac{-1}{-2} = -\frac{1}{2} \\
P = \frac{c}{a} = \frac{1}{-2} = -\frac{1}{2} \\
r_1 = -1,\quad r_2 = \frac{1}{2} \\
|r| > 1  \implies \text{invertible and stationary} \\
|r_1| = |-1| = 1 \\
|r_2| = |\frac{1}{2}| = \frac{1}{2} < 1
$$

8.AR(2)

$$
X_t = 2\beta X_{t-1} - \beta^2 X_{t-2} + Z_t \\
( 1-2\beta B + \beta^2 B^2 ) X_t = Z_t \\
( 1-2\beta B + \beta^2 B^2 ) = 0 \\
S = -\frac{b}{a} = -\frac{-2\beta}{\beta^2} = \frac{2}{\beta} \\
P = \frac{c}{a} = \frac{1}{\beta^2}  \\
r_1 = \frac{1}{\beta} = \beta^{-1} = r_2\\
|r| > 1  \implies \text{invertible and stationary} \\
|r| = |\beta^{-1}| = |\beta|^{-1}  > 1 \implies |\beta| < 1
$$

9.  AR(1), MA(2)

$$
X_t = 0.5 X_{t-1} + Z_t + 4 Z_{t-1} \\
(1-0.5B)X_t = (1 + 4B) Z_{t} \\
(1 + 4B) = 0 \implies r_{MA} = - \frac{1}{4}  \\
|r_{MA}| = | - \frac{1}{4} | = \frac{1}{4} < 1  \\
(1-0.5B) = 0 \implies r_{AR} = 2  \\
|r_{AR}| = | 2 | = 2 > 1 
$$

10. AR(1), MA(2)

$$
X_t = \beta^2 X_{t-1} + Z_t + 8\theta^3 Z_{t-1} \\
(1-\beta^2 B)X_t = (1 + 8\theta^3 B) Z_{t} \\
(1-\beta^2 B) = 0 \implies r_{MA} = \frac{1}{\beta^2}  \\
|r_{MA}| = | (\beta^2)^{-1} | = | \beta | ^{-2} \implies | \beta | < 1  \\
(1 + 8\theta^3 B) = 0 \implies r_{AR} = \frac{1}{-8\theta^3} = -(2\theta)^{-3} \\
|r_{AR}| = | -(2\theta)^{-3} | = | 2\theta | ^{-3} \implies | 2 \theta | < 1  \implies \theta < \frac{1}{2}
$$

## Yule Walker

TODO: check material from diff eq and Yule Walker

$$
a_n = \lambda^n
$$

$$
a_n=4a_{n-1}-3a_{n-2} \\
\lambda^2 - 4\lambda+3=0 \\
S = -\frac{b}{a} = -\frac{-4}{1} = 4 \\
P = \frac{c}{a} = \frac{3}{1} = 3 \\
\lambda_1 = 3, \quad \lambda_2=1
$$

$$
a_{n} = 4a_{n-1} - 3a_{n-2}  \\
a_n = c_1 + c_2 3^n \\
a_{2} = 4 a_1 - 3 a_0 = 4 (-2) - 3 (2) = -8 -6 = -14 \\
a_n = 4 - 2 (3^n) \\
a_0 = 4 - 2 (3^0) = 4 -2 = 2 \\
a_1 = 4 - 2 (3^1) = 4 - 6 = -2 \\
a_2 = 4 - 2 (3^2) = 4 - 18 = -14
$$

AR(1) starting at lag 1, even function

$$
X_t=0.4X_{t-1}+Z_t \\
\rho(k)=0.4\rho(k-1), \quad k \geq 1 \\
\rho(0) = 1 \\
\rho(k) = 0.4^{k}, \quad k \geq 0 \\
\rho(k) = \rho(-k) \quad\forall k \in \mathbb{Z}^{-}
$$

AR(3)

$$
X_t = \frac{1}{2} X_{t-1} + \frac{1}{9} X_{t-2} - \frac{1}{18} X_{t-3} + Z_{t} \\
18\lambda^3-9 \lambda^2 - 2 \lambda + 1 = 0 \\
(2\lambda-1)(3\lambda-1)(3\lambda+1)=0 \\
\rho(k) = \frac{1}{2} \rho(k-1) + \frac{1}{9} \rho(k-2)  - \frac{1}{18} \rho(k-3) \\
\rho(k) = c_1 \left( \frac{1}{2} \right)^k + c_2 \left( \frac{1}{3} \right)^k + c_3 \left( -\frac{1}{3} \right)^k
$$

# Week-3

MA(q): ACF cuts off after q lags

AR(p): PACF cuts off after p lags

The excess correlation at ð‘™ð‘Žð‘” = ð‘˜ not accounted for by a (ð‘˜ - 1)ð‘ ð‘¡ order
model, is the\
partial correlation at lag=k\

## PACF

```{r, fig.height=15, fig.width=15}
#{r}
#{r, fig.asp = .8}
# {r, echo=TRUE, fig.height=15, fig.width=15, out.retina=1}
phi = c(-.9, -.6)
par(mfrow=c(3,1))
data.ts = arima.sim(n = 500, list(ar = phi ))
plot(data.ts, 
main = 
paste(c("Autoregressive Process with phi=(", phi , ")"), collapse=" ") )
acf(data.ts)
acf(data.ts, type="partial")
```

```{r, fig.height=15, fig.width=15}
par(mfrow=c(3,1))
phi = c(.9, -.6, .3)
data.ts = arima.sim(n = 500, list(ar=phi) )
plot(data.ts, main= paste(c("Autoregressive Process with phi=(", phi , ")"), collapse=" ") ) 
acf(data.ts)
acf(data.ts, type="partial")
```

```{r, fig.height=15, fig.width=15}
phi = c(.6, -.6)
par(mfrow=c(2,1))
data.ts = arima.sim(n = 1000, list(ar = phi ))
plot(data.ts, 
main = 
paste(c("Autoregressive Process with phi=(", phi , ")"), collapse=" ") )
acf(data.ts, type="partial")
```

## Wheat price

```{r}
library(data.table) 
# Get the data
beveridge <- fread('https://people.sc.fsu.edu/~jburkardt/datasets/time_series/beveridge_wheat.txt') 

beveridge.ts = ts(beveridge[,2], start=1500)

# Check MA
mean(beveridge$V2[1:31])
mean(beveridge$V2[2:32])

beveridge.MA = filter(beveridge.ts, rep(1/31, 31), sides = 2)
data.frame(beveridge$V1, beveridge$V2, beveridge.MA)

plot( beveridge.ts, ylab="price", main="Beveridge Wheat Price Data")
lines(beveridge.MA, col="red")
```

```{r, fig.height=15, fig.width=15}
par(mfrow=c(3,1))
Y = beveridge.ts/beveridge.MA
plot( Y, ylab="scaled price", main="Transformed Beveridge Wheat Price Data")
acf(na.omit(Y),
main="Autocorrelation Function of Transformed Beveridge Data")
acf(na.omit(Y), type="partial",
main="Partial Autocorrelation Function of Transformed Beveridge Data")
```

```{r}
ar(na.omit(Y), order.max = 5)
```

## Body fat

```{r}
# install.packages("isdals")
library(isdals)
data(bodyfat)
attach(bodyfat)
cor( cbind(Fat, Triceps,Thigh,Midarm) )

pairs( cbind( Fat, Triceps, Thigh, Midarm) )

# Check most correlated variables
# (Fat,Thigh); (Fat,Triceps); (Triceps,Thigh)
Fat.hat = predict(lm(Fat~Thigh))
Triceps.hat = predict( lm(Triceps~Thigh) )
cor( (Fat- Fat.hat), (Triceps- Triceps.hat) ) #returns 0.1749822
# install.packages("ppcor")
library(ppcor)
pcor( cbind( Fat, Triceps, Thigh) )


Fat.hat = predict(lm(Fat~Thigh+Midarm))
Triceps.hat = predict( lm(Triceps~Thigh+Midarm) )
cor( (Fat- Fat.hat), (Triceps- Triceps.hat) ) #returns 0.33815

pcor( cbind( Fat, Triceps, Thigh, Midarm) )


```

## Partial Correlation

Remove Thigh and Midarm dependencies

$$
Triceps(\phi - \hat{\phi}(Thigh,Midarm)), \quad \phi = Fat \\
Fat(\tau - \hat{\tau}(Thigh,Midarm)), \quad \tau = Triceps \\ \\
Triceps \times Fat
$$

## Lake Huron

```{r, fig.height=15, fig.width=15}
par( mfrow=c(3,1) )
plot(LakeHuron)
acf(LakeHuron)
acf(LakeHuron, type="partial")
```

```{r, fig.height=15, fig.width=15}

# install.packages("pracma")
require(pracma)  # detrend

LakeHuron.diff = diff(LakeHuron)

par( mfrow=c(3,1) )
plot(LakeHuron.diff)
acf(LakeHuron.diff)$acf[1:3]
acf(LakeHuron.diff, type="partial")$acf
```

```{r}

R=matrix(1,2,2) # matrix of dimension 2 by 2, with entries all 1's.
r=NULL
r[1:2]=acf(LakeHuron.diff, plot=F)$acf[2:3]
R[1,2]=r[1] # only diagonal entries are edited
R[2,1]=r[1] # only diagonal entries are edited
R

b=matrix(r,nrow=2,ncol=1)
b

# solve(R,b) solves Rx=b, and gives x=R^(-1)b vector
phi.hat=NULL
phi.hat=solve(R,b)[,1]
phi.hat

#variance estimation using Yule-Walker Estimator
c0=acf(LakeHuron.diff, type='covariance', plot=F)$acf[1]
c0

var.hat=c0*(1-sum(phi.hat*r))
var.hat

# constant term in the model
phi0.hat=mean(my.data)*(1-sum(phi.hat))
phi0.hat

cat("Constant:", phi0.hat," Coefficients:", phi.hat, " and Variance:", var.hat, '\n')

sqrt(var.hat)

1-phi.hat[1]
phi.hat[1] - phi.hat[2]
```

## Attitude

```{r}
# package: datasets
attach(attitude);
rcl = cbind(rating, complaints, learning);
cor(rcl)

rating.hat = predict( lm( rating ~ learning) )
complaints.hat = predict( lm( complaints~learning) )

cor( (rating - rating.hat ), ( complaints - complaints.hat ) )

pcor(rcl)
```

## Recruitment

```{r}
library(astsa)
my.data=rec

# Plot rec 
plot(rec, main='Recruitment time series', col='blue', lwd=3)
```

```{r, fig.height=15, fig.width=15}
# subtract mean to get a time series with mean zero
ar.process=my.data-mean(my.data)

# ACF and PACF of the process
par(mfrow=c(2,1))
acf(ar.process, main='Recruitment', col='red', lwd=3)
pacf(ar.process, main='Recruitment', col='green', lwd=3)
```

```{r}
# order
p=2

# sample autocorreleation function r
r=NULL
r[1:p]=acf(ar.process, plot=F)$acf[2:(p+1)]
cat('r=',r,'\n')

# matrix R
R=matrix(1,p,p) # matrix of dimension 2 by 2, with entries all 1's.

# define non-diagonal entires of R
for(i in 1:p){
	for(j in 1:p){
		if(i!=j)
			R[i,j]=r[abs(i-j)]
		}
	}
R

# b-column vector on the right
b=NULL
b=matrix(r,p,1)# b- column vector with no entries
b

# solve(R,b) solves Rx=b, and gives x=R^(-1)b vector
phi.hat=NULL
phi.hat=solve(R,b)[,1]
phi.hat

#variance estimation using Yule-Walker Estimator
c0=acf(ar.process, type='covariance', plot=F)$acf[1]
c0

var.hat=c0*(1-sum(phi.hat*r))
var.hat

# constant term in the model
phi0.hat=mean(my.data)*(1-sum(phi.hat))
phi0.hat

cat("Constant:", phi0.hat," Coefficients:", phi.hat, " and Variance:", var.hat, '\n')
```

## Johnson & Johnson

```{r}
library(astsa)
# Time plot for Johnson&Johnson
plot(JohnsonJohnson, main='Johnson&Johnosn earnings per share', col='blue', lwd=3)
```

```{r, fig.height=15, fig.width=15}
# log-return of Johnson&Johnson
jj.log.return=diff(log(JohnsonJohnson))
jj.log.return.mean.zero=jj.log.return-mean(jj.log.return)

# Plots for log-returns
par(mfrow=c(3,1))
plot(jj.log.return.mean.zero, main='Log-return (mean zero) of Johnson&Johnosn earnings per share')
acf(jj.log.return.mean.zero, main='ACF')
pacf(jj.log.return.mean.zero, main='PACF')
```

```{r}
# Order
p=4

# sample autocorreleation function r
r=NULL
r[1:p]=acf(jj.log.return.mean.zero, plot=F)$acf[2:(p+1)]
r

# matrix R
R=matrix(1,p,p) # matrix of dimension 4 by 4, with entries all 1's.
# define non-diagonal entires of R
for(i in 1:p){
	for(j in 1:p){
		if(i!=j)
			R[i,j]=r[abs(i-j)]
		}
	}
R

# b-column vector on the right
b=matrix(r,p,1)# b- column vector with no entries
b

phi.hat=solve(R,b)[,1]
phi.hat

# Variance estimation using Yule-Walker Estimator
c0=acf(jj.log.return.mean.zero, type='covariance', plot=F)$acf[1]
c0
var.hat=c0*(1-sum(phi.hat*r))
var.hat

# Constant term in the model
phi0.hat=mean(jj.log.return)*(1-sum(phi.hat))
phi0.hat

cat("Constant:", phi0.hat," Coefficients:", phi.hat, " and Variance:", var.hat, '\n')
```

## Yule-Walker matrix

### AR(2)

x*t=phi1\*x*(t-1)+phi2\*x\_(t-2)+z_t z_t\~ N(0, sigma\^2)

```{r, fig.height=15, fig.width=15}
set.seed(2017)
sigma=4
phi=c(1/3,1/2)
n=10000

# simulated ar process
ar2.process=arima.sim(n, model=list(ar=phi), sd=sigma)
ar2.process[1:5]

par(mfrow=c(3,1))
plot(ar2.process, main='Simulated AR(2)')
acf(ar2.process, main='ACF')
pacf(ar2.process, main='PACF')
```

```{r}

# acf terms
r=NULL
r[1:2]=acf(ar2.process, plot=T)$acf[2:3]  # 1: gamma(0)=1
r

R=matrix(1,2,2) # matrix of dimension 2 by 2, with entries all 1's.
R[1,2]=r[1] # only diagonal entries are edited
R[2,1]=r[1] # only diagonal entries are edited
R

b=matrix(r,nrow=2,ncol=1)# b- column vector with no entries
b
# coefficients estimation
phi.hat=solve(R,b)
phi.hat
# variance estimation
c0=acf(ar2.process, type='covariance', plot=F)$acf[1]
var.hat=c0*(1-sum(phi.hat*r))
var.hat
```

### AR(3)

x_t=phi1\*x\_(t-1)+phi2\* x\_(t-2)+\\phi_3\*x\_(t-3)+z_t

z_t\~ N(0, sigma\^2)

```{r, fig.height=15, fig.width=15}
set.seed(2017)
sigma=4
phi=c(1/3,1/2,7/100)
n=100000

ar3.process=arima.sim(n, model=list(ar=phi), sd=sigma)

par(mfrow=c(3,1))
plot(ar3.process, main='Simulated AR(3)')
acf(ar3.process, main='ACF')
pacf(ar3.process, main='PACF')
```

```{r}

r=NULL
r[1:3]=acf(ar3.process, plot=T)$acf[2:4]
r

R=matrix(1,3,3) 
R[1,2]=r[1] 
R[1,3]=r[2]
R[2,1]=r[1]
R[2,3]=r[1]
R[3,1]=r[2]
R[3,2]=r[1]
R

# b-column vector on the right
b=matrix(,3,1)# b- column vector with no entries
b[1,1]=r[1]
b[2,1]=r[2]
b[3,1]=r[3]
b

# solve Rx=b and find phi's
phi.hat=solve(R,b)
phi.hat

# sigma estimation
c0=acf(ar3.process, type='covariance', plot=F)$acf[1]
var.hat=c0*(1-sum(phi.hat*r))
var.hat
```

\$\$

X_t = \phi\_0 + \phi\*1 X*{t-1} +* \phi\*2 X{t-2} + \phi\*3 X\*{t-3} +
Z_t \$\$

### AR(p)

$$
X_t = \phi_0 + \phi_1 X_{t-1} + \phi_2 X_{t-2} + \cdots + \phi_p X_{t-p} + Z_t
$$

$$
E[X_{t-k}] = \mu, \quad E[Z_t] = 0
$$

Expectation$$
\mu = \phi_0 + \phi_1 \mu + \phi_2 \mu + \cdots + \phi_p \mu + 0
$$

$$
X_t - \mu = \phi_1 (X_{t-1} - \mu) + \phi_2 (X_{t-2} - \mu) + \cdots + \phi_p (X_{t-p} - \mu) + Z_t
$$

$$
\tilde{X}_{t-k} = X_{t-k} - \mu, \quad E[\tilde{X}_t] = E[X_t] - \mu = \mu - \mu = 0
$$

$$
\tilde{X}_t = \phi_1 \tilde{X}_{t-1} + \phi_2 \tilde{X}_{t-2} + \cdots + \phi_p \tilde{X}_{t-p} + Z_t
$$

AR(p) with null mean

$$
\mu = 0
$$

$$
X_t = \phi_1 X_{t-1} + \phi_2 X_{t-2} + \cdots + \phi_p X_{t-p} + Z_t
$$

$$
\rho(k) = \begin{cases}
\phi_1\rho(k-1)+\phi_2\rho(k-2)+\cdots+\phi_p\rho(k-p), & k\geq 1\\
1 & k=0 \\
\rho(-k) & k < 0
\end{cases}
$$

$$
\rho(1) = \phi_1\rho(0)+\phi_2\rho(-1)+\cdots+\phi_p\rho(1-p) \\
\rho(2) = \phi_1\rho(1)+\phi_2\rho(0)+\cdots+\phi_p\rho(2-p) \\
\vdots\\
\rho(p-1) = \phi_1\rho(p-2)+\phi_2\rho(p-3)+\cdots+\phi_p\rho(1) \\
\rho(p) = \phi_1\rho(p-1)+\phi_2\rho(p-2)+\cdots+\phi_p\rho(0)
$$

$$
\rho(1) = \phi_1+\phi_2\rho(1)+\cdots+\phi_p\rho(p-1) \\
\rho(2) = \phi_1\rho(1)+\phi_2+\cdots+\phi_p\rho(p-2) \\
\vdots\\
\rho(p-1) = \phi_1\rho(p-2)+\phi_2\rho(p-3)+\cdots+\phi_p\rho(1) \\
\rho(p) = \phi_1\rho(p-1)+\phi_2\rho(p-2)+\cdots+\phi_p
$$

$$
\begin{Bmatrix}
\rho(1) \\
\rho(2) \\
\vdots \\
\rho(p-1) \\
\rho(p)
\end{Bmatrix}
= 
\begin{bmatrix}
1 & \rho(1) & \rho(2) & \cdots & \rho(p-1) \\
\rho(1) & 1 & \rho(1) & \cdots & \rho(p-2) \\
\vdots & \vdots & \vdots & \vdots & \vdots \\
\rho(p-2) & \rho(p-3) & \rho(p-4) & \cdots & \rho(1) \\
\rho(p-1) & \rho(p-2) & \rho(p-3) &  \cdots & 1 \\
\end{bmatrix}
\begin{Bmatrix}
\phi(1) \\
\phi(2) \\
\vdots \\
\phi(p-1) \\
\phi(p)
\end{Bmatrix}\\
b = R \phi \quad \text{unique solution}\\
R^{-1} b = \phi
$$

$R$: symmetric, positive semidefinite, non-negative eigenvalues,
invertible

Discrete:
$r\approx\rho,\quad \hat{\phi} \approx \phi,\quad \hat{R} \approx R$

AR(2), with ($\phi_0=0$)

$$
X_t = \phi_1 X_{t-1} + \phi_2 X_{t-2} + Z_t
$$

$$
\begin{Bmatrix}
r_1 \\
r_2
\end{Bmatrix}
= 
\begin{bmatrix}
1 & r_1 \\
r_1 & 1
\end{bmatrix}
\begin{Bmatrix}
\hat{\phi}_1 \\
\hat{\phi}_2 
\end{Bmatrix}
$$

$$
r_1 = \hat{\phi}_1 + r_1 \hat{\phi}_2 \\
r_2 = r_1 \hat{\phi}_1 + \hat{\phi}_2
$$

$$
Z\sim\mathcal{N}(E[Z_t]=0,V[Z_t]=\sigma_Z^2)
$$

$$
E[X_t] = \mu
$$

$$
E[X_t] = \phi_1 E[ X_{t-1} ] +  \phi_2 E[ X_{t-2} ] + E [Z_t] \\
\mu = \phi_1 \mu +  \phi_2 \mu + 0 = \mu (\phi_1+\phi_2) \implies \mu=0
$$

$$
E[X_{t-k}X_t] = \phi_1 E[ X_{t-k} X_{t-1} ] +  \phi_2 E[ X_{t-k} X_{t-2} ] + E [ X_{t-k} Z_t] \\
$$

$$
\mathrm{COV}[X_t, X_{t+k}] = \mathrm{E}\left[ X_{t} \cdot X_{t+k} \right] - \mathrm{E}\left[ X_{t} \right] \mathrm{E}\left[  X_{t+k} \right]
$$

$$
\gamma(s,t) = \mathrm{Cov}[X_s,X_t] = \mathrm{E}\left[ (X_s - \mu_s) (X_t - \mu_t) \right] 
$$

$$
V[X_t] = \phi_1^2 V[ X_{t-1} ] +  \phi_2^2 V[ X_{t-2} ] 
+ 2 \phi_1 \phi_2 cov(X_{t-1}, X_{t-2}) + V[Z_t]
$$

TODO: VER ESSE PASSO (COMO SURGE GAMMA 0 e GAMMA 1)

$$
V[X_t] = V[X_{t-1}] = V[X_{t-2}] = \gamma(0) \\
cov(X_{t-1}, X_{t-2}) = \frac{\gamma(1)}{\gamma(0)} = \rho_1
$$

$$
\sigma_Z^2 = \gamma(0) \left[1-\phi_1^2 - \phi_2^2 -
 2 \phi_1 \phi_2 \rho_1 \right]
$$

$$
1-\phi_1^2 - \phi_2^2 -  2 \phi_1 \phi_2 \rho_1 \\
= 1-\phi_1^2 - \phi_1 \phi_2 \rho_1 - \phi_2^2 -  \phi_1 \phi_2 \rho_1 \\
= 1-\phi_1(\phi_1 + \rho_1\phi_2)  - \phi_2(\phi_2 +  \phi_1 \rho_1 )
$$

From the matrix equation:

$$
\rho_1 = \phi_1 + \rho_1\phi_2, \quad \rho_2 = \phi_2 + \rho_1\phi_1
$$

Therefore the variance:

$$
\sigma_Z^2 = \gamma(0) \left[1-\phi_1\rho_1 - \phi_2\rho_2  \right] \\
\hat{\sigma}_Z^2 = c_0 \left[1-\hat{\phi}_1 r_1 - \hat{\phi}_2 r_2  \right] = c_0 \left[1- \sum_{i=1}^{p} \hat{\phi}_i r_i  \right]
$$

AR(3), with ($\phi_0=0$)

$$
X_t = \phi_1 X_{t-1} + \phi_2 X_{t-2} + \phi_3 X_{t-3} + Z_t
$$

$$
\begin{Bmatrix}
r_1 \\
r_2 \\
r_3
\end{Bmatrix}
= 
\begin{bmatrix}
1 & r_1 & r_2 \\
r_1 & 1 & r_1 \\
r_2 & r_1 & 1
\end{bmatrix}
\begin{Bmatrix}
\hat{\phi}_1 \\
\hat{\phi}_2 \\
\hat{\phi}_3
\end{Bmatrix}
$$

```{r}
r=c(0.8,0.6,0.2)

R=matrix(1,3,3) 
R[1,2]=r[1] 
R[1,3]=r[2]
R[2,1]=r[1]
R[2,3]=r[1]
R[3,1]=r[2]
R[3,2]=r[1]
R

# b-column vector on the right
b=matrix(,3,1)# b- column vector with no entries
b[1,1]=r[1]
b[2,1]=r[2]
b[3,1]=r[3]
b

# solve Rx=b and find phi's
phi.hat=solve(R,b)
phi.hat

# sigma estimation
c0=5
var.hat=c0*(1-sum(phi.hat*r))
var.hat
```

# Week-4

## AIC

-   credit: reduce the error sum of squares

-   penalty: add parameters

$$
ð´ð¼ð¶ = -2 \log(MLE) + 2p 
$$

$$
ð´ð¼ð¶ = \log(\hat{\sigma}^2) + \frac{n + 2p}{n}, \quad \hat{\sigma}^2 = \frac{SSE}{n}
$$

-   $MLE$ : maximum likelihood

-   $p$: number of parameters in the model

-   $n$: number of samples

-   $SSE$: error sum of squares

AR(2)

$$
X_t = 0.7X_{t-1}-0.2X_{t-2}+Z_t
$$

```{r}
#{r, fig.height=15, fig.width=15}
rm(list=ls(all=TRUE))
set.seed(43) #Roman conquest of Britain
data = arima.sim( list(order = c(2,0,0), ar =c( 0.7, -.2)), n = 2000)

par(mfrow=c(1,2))
acf(data, main="ACF of AR Data of Second Order")
acf(data, type="partial", main="PACF of Time Series")

```

```{r}
arima(data, order=c(2,0,0), include.mean=FALSE )
```

```{r}
SSE=NULL
AIC=NULL
for (p in 1:5) {
m = arima(data, order=c(p,0,0), include.mean=FALSE )
SSE[p] = sum(resid(m)^2)
AIC[p] = m$aic
print( m$coef )
print( paste(m$aic, sum(resid(m)^2)) )
}


```

```{r}
par(mfrow=c(1,2))
order=c(1,2,3,4,5)
plot(SSE~order, main="SSE plotted on Order of AR(p) Model", ylim=c(1800, 2100))
plot(AIC~order, main="AIC plotted on Order of AR(p) Model", ylim=c(5500, 5800))
```

```{r}
rm(list=ls(all=TRUE))
set.seed(500) #500	Seven kingdoms Kent, Essex, Sussex, Wessex, East Anglia, Mercia, and Northumbria.)
data = arima.sim( list(order = c(3,0,0), ar =c( 0.6, -0.1, .4)), n = 5000)

arima(data, order=c(2,0,0), include.mean=FALSE )
arima(data, order=c(3,0,0), include.mean=FALSE )
arima(data, order=c(4,0,0), include.mean=FALSE )
```

```{r}
rm(list=ls(all=TRUE))
set.seed(597) # Saint Augustine arrives in England
data = arima.sim( list(order = c(1,0,0), ar = .3), n = 5000)

par(mfrow=c(1,2))
acf(data, main="ACF of Time Series Data")
acf(data, type="partial", main="PACF of Time Series Data")
```

```{r}
arima(data, order=c(1,0,0) );
arima(data, order=c(2,0,0) );
arima(data, order=c(3,0,0) );
```

```{r}
library(forecast)
data = arima.sim( n=1E4, list(ar=.5, ma=.2) )
auto.arima(data)
```


## ARMA

ARMA(p,q)
$$
X_t = \phi_1 X_{t-1} + \cdots + \phi_p X_{t-p} + \beta_0 Z_t + \beta_1 Z_{t-1} + \cdots + \beta_q Z_{t-q} \\
\phi(B)X_t = \beta(B)Z_t \\
\phi(B) = 1 - \phi_1 B - \cdots - \phi_p B^{p} \\
\beta(B) = \beta_0 + \beta_1 B - \cdots - \beta_q B^{q}
$$
$$
X_t = \sum_{i=1}^{n=p} \phi_i X_{t-i} + \sum_{i=0}^{n=q} \theta_i Z_{t-i}
$$

Terms
* Noise: $Z_t$
* AR: $\sum_{i=1}^{n=p} \phi_i X_{t-i}$
* MA: $\sum_{i=1}^{n=q} \theta_i Z_{t-i}$

Operators
$$
\Theta(B)Z_t = \Phi(B) X_t \\
\Psi(B) = \frac{\Theta(B)}{\Phi(B)}
$$

$$
X_t  = \frac{\Theta(B)}{\Phi(B)}Z_t = \Psi(B) Z_t
$$


MA mixed
$$
X_t = 0.5X_{t-1} + Z_t + 0.2 Z_{t-1} \\
(1-0.5B)X_t = (1+0.2B) Z_t \\
X_t = \frac{(1+0.2B)}{(1-0.5B)}Z_t \\
X_t = (1+0.2B)(1+0.5B+0.25B^2+0.125B^3 + \cdots)Z_t \\
X_t = (1+0.5B+0.25B^2+0.125B^3 + 0.2B + 0.1B^2+0.05B^3+0.025B^4 + \cdots)Z_t \\
X_t = (1+0.7B+0.35B^2+0.175B^3 + \cdots)Z_t \\
$$

The coefficients on the polynomial should decay rather quickly.

```{r}
phi=.5
beta=0.2
sigma=3
m=10000

set.seed(5)
Simulated.Arima = 
    arima.sim(
      n=m,
      list(order = c(1,0,1), 
           ar = phi, 
           ma=beta)
      )
acf(Simulated.Arima)
```



## ARIMA

```{r}
# parameters
phi=c(.7, .2)
beta=0.5
sigma=3
m=10000

set.seed(5)
(Simulated.Arima = 
    arima.sim(
      n=m,
      list(order = c(2,1,1), 
           ar = phi, 
           ma=beta)
      )
  )
```

```{r}
plot(Simulated.Arima, ylab=' ',main='Simulated time series from ARIMA(2,1,1) process', col='blue', lwd=2)
```

```{r}
acf(Simulated.Arima)
```

```{r}
Diff.Simulated.Arima=diff(Simulated.Arima)
plot(Diff.Simulated.Arima)
```

```{r}
acf(Diff.Simulated.Arima)
```

```{r}
pacf(Diff.Simulated.Arima)
```

```{r}
library(astsa)
sarima(Simulated.Arima,2,1,1,0,0,0) # astsa not installed
```

```{r}
library(forecast)
auto.arima(Simulated.Arima) # forecast not installed
```

```{r}
(fit1<-arima(Diff.Simulated.Arima, order=c(4,0,0)))
```

```{r}
(fit2<-arima(Diff.Simulated.Arima, order=c(2,0,1)))
```

```{r}
(fit3<-arima(Simulated.Arima, order=c(2,1,1)))
```


## Daily female births in California, 1959

The following time series is taken from Time Series Data Library (TSDL)
TSDL was created by Rob Hyndman
Professor of Statistics at Monash University, Australia.
====== Daily total female birth in California, 1959 =======
Data is exported as csv file to the working directory
Link: https://datamarket.com/data/list/?q=cat:fwy%20provider:tsdl



```{r}
library(astsa)

# read data to R variable
birth.data<-read.csv("./data/daily-total-female-births-in-cal.csv")

# pull out number of births column
number_of_births<-birth.data$Daily.total.female.births.in.California..1959

# use date format for dates
birth.data$Date <- as.Date(birth.data$Date, "%m/%d/%Y")

plot.ts(number_of_births,main='Daily total female births in california, 1959', ylab = 'Number of births')
```

```{r}
# Test for correlation
Box.test(number_of_births, lag = log(length(number_of_births)))
```

```{r}
# Plot the differenced data
plot.ts(diff(number_of_births), main='Differenced series', ylab = '')
```


```{r}
# Test for correlation in the differenced data
Box.test(diff(number_of_births), lag = log(length(diff(number_of_births))))
```

```{r}
# acf of the differenced data
acf(diff(number_of_births), main='ACF of differenced data', 50)

```
```{r}
# pacf of the differenced data
pacf(diff(number_of_births), main='PACF of differenced data', 50)
```

```{r}
# Fit various ARIMA models


model1<-arima(number_of_births, order=c(0,1,1))
SSE1<-sum(model1$residuals^2)
model1.test<-Box.test(model1$residuals, lag = log(length(model1$residuals)))

model2<-arima(number_of_births, order=c(0,1,2))
SSE2<-sum(model2$residuals^2)
model2.test<-Box.test(model2$residuals, lag = log(length(model2$residuals)))

model3<-arima(number_of_births, order=c(7,1,1))
SSE3<-sum(model3$residuals^2)
model3.test<-Box.test(model3$residuals, lag = log(length(model3$residuals)))

model4<-arima(number_of_births, order=c(7,1,2))
SSE4<-sum(model4$residuals^2)
model4.test<-Box.test(model4$residuals, lag = log(length(model4$residuals)))

df<-data.frame(row.names=c('AIC', 'SSE', 'p-value'), c(model1$aic, SSE1, model1.test$p.value), 
               c(model2$aic, SSE2, model2.test$p.value), c(model3$aic, SSE3, model3.test$p.value),
               c(model4$aic, SSE4, model4.test$p.value))
colnames(df)<-c('Arima(0,1,1)','Arima(0,1,2)', 'Arima(7,1,1)', 'Arima(7,1,2)')



format(df, scientific=FALSE)
```

```{r}
# Fit a SARIMA model

sarima(number_of_births, 0,1,2,0,0,0)

```



## Ljung-Box Q Statistic


ARMA(p,q)
$$
X_t = \phi_1 X_{t-1} + \cdots + \phi_p X_{t-p} + \beta_0 Z_t + \beta_1 Z_{t-1} + \cdots + \beta_q Z_{t-q} \\
\phi(B)X_t = \beta(B)Z_t \\
\phi(B) = 1 - \phi_1 B - \cdots - \phi_p B^{p} \\
\beta(B) = \beta_0 + \beta_1 B - \cdots - \beta_q B^{q}
$$

Difference operator
* remove trend from non-stationary processes

$$
\nabla = 1 -B
$$


$$
\nabla X_t = X_t - X_{t-1} = (1-B)X_t
$$


Random walk
$$
X_t = X_{t-1} + Z_t \\
\nabla X_t = Z_t
$$

ARIMA(p,d,q)
Autorregressive Integrated Moving Average of order (p,d,q)
$$
Y_t := \nabla^d X_t = (1-B)^d X_t \\
Y_t \sim \text{ARMA}(p,q)
\implies
X_t \sim \text{ARIMA}(p,d,q)
$$


$$
\phi(B)\nabla^d X_t = \beta(B) Z_t \\
\phi(B) ( 1-B) ^d X_t = \beta(B) Z_t
$$


Order of differencing $d$
* $d=1$ or $d=2$
* over differencing may introduce dependence
* ACF might also suggest differencing is needed
* $\phi(z)(1-z)^d$ has unit root with multiplicity $d$
* ACF will decay very slowly


ARIMA(3,0,2)
$$
X_t = \phi_1 X_{t-1} + \phi_2 X_{t-2} + \phi_3 X_{t-3} + Z_t + \beta_1 Z_{t-1} +  \beta_2 Z_{t-2} \\
(1-\phi_1 B - \phi_2 B^2 - \phi_3 B^3 )X_t = (1-\beta_1 B - \beta_2 B^2)  Z_t
$$

$$
(1+0.2B)\nabla X_t = (1-0.3B)Z_t \\
(1+0.2B)(1-B) X_t = (1-0.3B)Z_t \\
(1+0.2B-B-0.2B^2) X_t = (1-0.3B)Z_t \\
(1-0.8B-0.2B^2) X_t = (1-0.3B)Z_t \\
X_t = 0.8X_{t-1} + 0.2 X_{t-2} + Z_t - 0.3Z_{t-1}
$$




$$
X_t = 3 X_{t-2} - 2 X_{t-3} + Z_{t} + 5 Z_{t-1}  \\
( 1 - 3 B^2 + 2 B^3 )X_t =  (1-5B) Z_{t}  \\
$$

ARIMA(2,1,1)
* Since $1-3 \cdot 1^2+2 \cdot 1^3=0$, we obtain that $1-B$ divides $1-3B^2+2B^3$.
* Long division would give us $1-3B^2+2B^3=(1+B-2B^2)(1-B)$.
* Here $(1+B-2B^2)(1+Bâˆ’2B^2)$ could be taken as AR polynomial.
* Note that the factor $(1-B)$ would make the process nonstationary. 

ARIMA(3,0,1)
* Even though there are 2 AR terms, the order is 3 since we have $-2X_{t-3}$
* And there is 1 MA term.
* Note that the process is not invertible since the root of the polynomial $1+5B$ lies inside the unit circle. How about stationarity?

ARIMA(1,2,1)
* $1 âˆ’3B^2 +2B^3$ can be factorized as $(1+2B)(1-B)^2$,
where we have 1 AR terms (degree of the AR polynomial $1+2B$), and two differencing $(1-B)^2$. Note that process is nonstationary. Why is that? 


We have some time series whose Q-statistic at lag=4 is calculated, and corresponding p-value is found: p-value=0.340.34. What does it mean

* We do not have enough evidence to reject the null hypothesis that there is no autocorrelation UNTIL lag 4.
* The null hypothesis is $H_0: \rho_1=\rho_2=\rho_3=\rho_4=0$
* Since 0.34 is not smaller than any reasonable significance level, we cannot reject the null hypothesis.


We have some time series whose Q-statistic at lag=10 is calculated, and corresponding p-value is found: p-value=0.00034. What does it mean?

* There is a significant autocorrelation at at least one lag until lag 10.
* A low p-value is an evidence against the null hypothesis.
* We do have sufficient evidence to reject that all autocorrelation coefficients until lag 10 is zero. 
* p-value if the probability of obtained data given that the null hypothesis is correct. A low p-value is an evidence against the null hypothesis.

## BJsales

```{r}
BJsales
plot(BJsales)
```

* There are ups and downs with a general upward trend.  
* Time series is not stationary.
  * There is a trend which means mean is changing. 

```{r}
plot(diff(BJsales))
```
* It does not seem to be stationary since there are still upward or downward trends in different parts of the time plot.
 * There is, for example, an upward trend between (50,100) but a downward trend between (100,150). That means the mean is changing.  

```{r}
plot(diff(diff(BJsales)))
```

* There is no systematic change in mean. 
 * Mean level seems to be constant, around 0.
* Variance towards the end of the series seems to be different from the variance in the other parts of the plot. 
 * It seems that variance is smaller towards the end of the plot. One may say that difference in the change of the variance is not high, and thus can be ignored. 

```{r}
(pacf(diff(diff(BJsales))))
```

* Lags: 1, 2, 3, 10, 19
* Keeping parsimony principle in mind, the order of AR terms can be 0,1,2 or 3.
* One might say that Lag 19 is barely significant.
* AR(p): PACF cuts off after p lags







```{r}
(acf(diff(diff(BJsales))))
```
* Lags: 1, 8, 11
* If we ignore barely significant lags, the order of MA term can be 0 or 1.
* Lag  8 and Lag 11 are barely significant.
* MA(q): ACF cuts off after q lags
* Keeping parsimony principle in mind, the order of MA term can be 0 or 1.



```{r}
d=2
for(p in 1:4){
  for(q in 1:2){
        if(p+d+q<=8){
          model<-arima(x=BJsales, order = c((p-1),d,(q-1)))
          pval<-Box.test(model$residuals, lag=log(length(model$residuals)))
          sse<-sum(model$residuals^2)
          cat(p-1,d,q-1, 'AIC=', model$aic, ' SSE=',sse,' p-VALUE=', pval$p.value,'\n')
        }
      }
}
```


* AIC: ARIMA(0,2,1), Smallest AIC=517.1371
* SSE: ARIMA(3,2,1), Smallest SSE value is 264.0684.

```{r, fig.height=12, fig.width=16}
model<-arima(BJsales, order=c(0,2,1))

par(mfrow=c(2,2))

plot(model$residuals)
acf(model$residuals)
pacf(model$residuals)
qqnorm(model$residuals)
```

Is there compelling evidence against the whiteness of the residuals?
* No, since QQ-plot seems linear.
* No, since ACF and PACF has no significant lags. 

```{r}
summary(model)
```

$$
Y_t = \nabla^2 X_t = (1+\beta_1 B)Z_t = Z_t +\beta_1 Z_{t-1} \\
(1-B)^2 X_t = (1-2B+B^2) X_t = (1+\beta_1 B)Z_t \\
X_t = 2X_{t-1} - X_{t-2} +  Z_t +\beta_1 Z_{t-1} \\
X_t -2X_{t-1} + X_{t-2} = Z_t +\beta_1 Z_{t-1}
$$


# Week-5


## SARIMA(p,d,q,P,D,Q)S
* p: $X_t$
* d
* q: $Z_t$
* P: $X_t$
* D
* Q: $Z_t$
* S: 12



S=12
$$
(1+0.1B-0.2B^2)(1+0.25B^{12})X_t
= (1-0.3B)(1+0.5B^{12})Z_t\\
(1+0.1B-0.2B^2) \implies p=2 \\
(1+0.25B^{12}) \implies P=1 \\
(1-0.3B) \implies q=1 \\
(1+0.5B^{12}) \implies Q=1 \\
SARIMA(2,0,1,1,0,1)_{12} \\
(1+0.1B-0.2B^2+0.25B^{12}+0.025B^{13}-0.05B^{14})X_t
= (1+0.5B^{12}-0.3B-0.15B^{13})Z_t\\
\implies SARIMA(14,0,13,0,0,0)_1
$$

There is no differencing in the process, 
thus $d=0$ and $D=0$. 
AR polynomial gives $p=2$, 
and seasonal AR polynomial with degree $12*1$ gives $P=1$.

The total number of parameters in this model is 27+1=28. 
In modeling, we prefer the simplest possible model to avoid overfitting. 
Including the span of the seasonality of 12 would give us a simpler model.



S=24
$$
(1+0.1B)(1+0.25B^{12}-0.7B^{24})(1-B)(1-B^{12})X_t
= Z_t\\
(1+0.1B) \implies p=1 \\
(1+0.25B^{12}-0.7B^{24}) \implies P=2 \\
(1-B) \implies d=1 \\
(1-B^{12}) \implies D=1 \\
SARIMA(1,1,0,2,1,0)_{12} \\
Z_t = (1+0.25B^{12}-0.7B^{24}+0.1B+0.025B^{13}-0.07B^{25})(1-B-B^{12}+B^{13})X_t 
\\
Z_t = (1+0.25B^{12}-0.7B^{24}+0.1B+0.025B^{13}-0.07B^{25} \\
-B-0.25B^{13}+0.7B^{25}-0.1B^{2}-0.025B^{14}+0.07B^{26} \\
-B^{11}-0.25B^{24}+0.7B^{36}-0.1B^{13}-0.025B^{25}+0.07B^{37} \\
B^{12}+0.25B^{25}-0.7B^{37}+0.1B^{14}+0.025B^{26}-0.07B^{38})X_t \\
Z_t = (1-0.9B-0.1B^{2} \\
-B^{11}+1.25B^{12}-0.325B^{13}+0.0975B^{14} \\
-0.95B^{24}+0.855B^{25}+0.095B^{26} \\
+0.7B^{36}-0.63B^{37}-0.07B^{38})X_t \\
\implies SARIMA(38,0,0,0,0,0)_{1}
$$

Seasonal AR polynomial has a degree of $24=2*12$ which gives us $P=2$.

Probably useless, but correct. 
If we expand the polynomial $(1+0.1 B)(1+0.25 B^{12}-0.7 B^{48})(1-B)(1-B^{12})$,
we obtain an AR polynomial with a degree of $38$.


$SARIMA(0,0,2,0,0,1)_{12}$

MA: 0.2, 0.3
SMA: 0.5
$\sigma^2_Z=1$


$$
X_t = (1+0.2B+0.3B^2)(1+0.5B^{12})Z_t \\
X_t = (1+0.2B+0.3B^2+0.5B^{12}+0.1B^{13}+0.15B^{14})Z_t \\
X_t = Z_t+0.2Z_{t-1}+0.3Z_{t-2}+0.5Z_{t-12}+0.1Z_{t-13}+0.15Z_{t-14}
$$


TODO: check why
$$
\rho(3)=0
$$


Check the name of lambda
$$
X_{t-10} = Z_{t-10}+0.2Z_{t-11}+0.3Z_{t-12}+0.5Z_{t-22}+0.1Z_{t-23}+0.15Z_{t-24} \\
Z_{t-12} \implies Cov(X_t,X_{t-10}) = 0.3\cdot 0.5 = 0.15 \implies \lambda(10) = 0.15
$$
$Z_{tâˆ’12}$ is common term with coefficient 0.3 and 0.5 for both equations 
of $X_t$ and $X_{t-10}$.



## Fit SARIMA Process

1. Time plot
2. Transformation (stationary)
3. Differencing (seasonal or non-seasonal)
4. ACF: Closers spikes: MA order
5. ACF: spikes around seasonal lags: SMA order
6. PACF: Closers spikes: AR order
7. PACF: spikes around seasonal lags: SAR order
8. Fit few different models
9. Compare AIC, SSE, choose model with minimum AIC
10. Parsimony principle (order sum <= 6)
11. Time plot, ACF, PACF of residuals
12. Ljung-Box test for residuals

## Examples
```{r}
x=NULL
z=NULL
n=10000

z=rnorm(n)

# TODO: CHECK WHY THIS
x[1:13]=1

for(i in 14:n){
  x[i]<-z[i]+0.7*z[i-1]+0.6*z[i-12]+0.42*z[i-13]
}

plot.ts(x[12:120], main='The first 10 months of simulation SARIMA(0,0,1,0,0,1)_12', ylab='') 


```

```{r}
acf(x, main='SARIMA(0,0,1,0,0,1)_12 Simulation')
```




## Johnson e Johnson


```{r}
library(astsa)

d=1
DD=1

per=4

for(p in 1:2){
  for(q in 1:2){
    for(i in 1:2){
      for(j in 1:2){
        if(p+d+q+i+DD+j<=10){
          model<-arima(x=log(jj), order = c((p-1),d,(q-1)), seasonal = list(order=c((i-1),DD,(j-1)), period=per))
          pval<-Box.test(model$residuals, lag=log(length(model$residuals)))
          sse<-sum(model$residuals^2)
          cat(p-1,d,q-1,i-1,DD,j-1,per, 'AIC=', model$aic, ' SSE=',sse,' p-VALUE=', pval$p.value,'\n')
        }
      }
    }
  }
}
```


```{r}
sarima(log(jj), 0,1,1,1,1,0,4)
```

## Milk production

The time series is downloaded from TSDL.

```{r}
milk<-read.csv('./data/monthly-milk-production-pounds-p.csv')
Milk<-milk$Pounds
plot(Milk)
```

```{r}
par(mar=c(3,3,3,1))
acf2(Milk)
```


```{r}
plot(diff(Milk))
plot(diff(diff(Milk)))
diff2 = diff(diff(Milk),12)
plot(diff2)
```

```{r}
par(mar=c(3,3,3,1))
acf(diff2, lag.max=50, main='ACF')
par(mar=c(3,3,3,1))
pacf(diff2, lag.max=50, main='PACF')
```

diff2: make the process stationary
ACF:  q=0; Q=0,1,2,3 (seasonal at 12, 24, 36)
PACF: p=0; P=0,1,2



```{r}
library(astsa)
library(forecast)

d=NULL
DD=NULL
d=1
DD=1

per=12
for(p in 1:1){
  for(q in 1:1){
    for(i in 1:3){
      for(j in 1:4){
        if(p+d+q+i+DD+j<=10){
          model<-arima(x=Milk, order = c((p-1),d,(q-1)), seasonal = list(order=c((i-1),DD,(j-1)), period=per))
          pval<-Box.test(model$residuals, lag=log(length(model$residuals)))
          sse<-sum(model$residuals^2)
          cat(p-1,d,q-1,i-1,DD,j-1,per, 'AIC=', model$aic, ' SSE=',sse,' p-VALUE=', pval$p.value,'\n')
        }
      }
    }
  }
}
```

```{r}
library(astsa)
sarima(Milk, 0,1,0,0,1,1,12)
```


$$
(1-B)(1-B^{12})X_t = (1+\Theta B^{12})Z_t \\
X_t = X_{t-1} + X_{t-12} - X_{t-13} + Z_t + \Theta Z_{t-12} \\
\hat{\Theta} = - 0.675, \quad |_t \sim \mathcal{N} (\mu=0, \sigma^2=34.47)
$$


```{r}
model<- arima(x=Milk, order = c(0,1,0), seasonal = list(order=c(0,1,1), period=12))
summary(model)
```
```{r}
forecast(model)
```

```{r}
plot(forecast(model))
```

## Sales at a souvenir shop

The time series is downloaded from TSDL.
https://datamarket.com/data/set/22mh/monthly-sales-for-a-souvenir-shop-on-the-wharf-at-a-beach-resort-town-in-queensland-australia-jan-1987-dec-1993#!ds=22mh&display=line


```{r, fig.width=12, fig.height=8}
SUV<-read.csv('./data/monthly-sales-for-a-souvenir-sho.csv')
suv<-ts(SUV$Sales)

library(astsa)
library(forecast)



par(mfrow=c(2,2))

plot(suv, main='Monthly sales for a souvenir shop', ylab='', col='blue', lwd=3)
plot(log(suv), main='Log-transform of sales', ylab='', col='red', lwd=3)
plot(diff(log(suv)), main='Differenced Log-transform of sales', ylab='', col='brown', lwd=3)
plot(diff(diff(log(suv)),12), main='Log-transform without trend and seasonality', ylab='', col='green', lwd=3)
```

Assume that seasonal diff is stationary. Actually there is a variance variation.


```{r}
par(mar=c(3,3,3,2))
acf2(suv)
```



```{r}
data<-diff(diff((log(suv)),12))
par(mar=c(3,3,3,2))
acf2(data, 50)
```

ACF: q=0,1; Q=0,1,2,3
PACF: p=0,1; P=0,1


```{r}
d=1
DD=1
per=12
for(p in 1:2){
  for(q in 1:2){
    for(i in 1:2){
      for(j in 1:4){
        if(p+d+q+i+DD+j<=10){
          model<-arima(x=log(suv), order = c((p-1),d,(q-1)), seasonal = list(order=c((i-1),DD,(j-1)), period=per))
          pval<-Box.test(model$residuals, lag=log(length(model$residuals)))
          sse<-sum(model$residuals^2)
          cat(p-1,d,q-1,i-1,DD,j-1,per, 'AIC=', model$aic, ' SSE=',sse,' p-VALUE=', pval$p.value,'\n')
        }
      }
    }
  }
}
```


```{r}
model<- arima(x=log(suv), order = c(1,1,0), seasonal = list(order=c(0,1,1), period=12))
summary(model)
```


$$
Y_t = \log(X_t) \\
(1-\phi B)(1-B)(1-B^{12})Y_t = (1+\Theta B^{12})Z_t \\
Y_t = (1+\phi) Y_{t-1} - phi Y_{t-2} - (1+\phi) Y_{t-13} + \phi Y_{t-14} + Z_t + \Theta Z_{t-12} \\
\hat{\phi} = - 0.5017, \quad  \hat{\Theta} = - 0.5107, \quad |_t \sim \mathcal{N} (\mu=0, \sigma^2=0.03111)
$$


```{r}
sarima(log(suv),1,1,0,0,1,1,12)
```

Residuals analysis
standardized residual looks white
no autocorrelation
qqplot: linear, systematic departure at tails
p-values no significance correlation in the residuals



```{r}
forecast(model)
```

```{r}
plot(forecast(model))
```

```{r}
a<-sarima.for(log(suv),12,1,1,0,0,1,1,12)

plot.ts(c(suv,exp(a$pred)), main='Monthly sales + Forecast', ylab='', col='blue', lwd=3)
```


```{r}
sarima(log(suv),0,1,3,0,1,1,12)
```



## USAccDeaths


1. Time plot
2. Transformation (stationary)
3. Differencing (seasonal or non-seasonal)


```{r, fig.width=12, fig.height=8}

data = USAccDeaths

library(astsa)
library(forecast)

par(mfrow=c(2,2))

plot(data, main='', ylab='', col='blue', lwd=3)
plot(log(data), main='Log-transform', ylab='', col='red', lwd=3)
plot(diff(log(data)), main='Differenced Log-transform', ylab='', col='brown', lwd=3)
plot(diff(diff(log(data)),12), main='Log-transform without trend and seasonality', ylab='', col='green', lwd=3)
```

```{r}
help(USAacDeaths)
```

It is a monthly time series with a span of seasonality 12. 
help(USAacDeaths) routine gives more information about the series including being monthly time series.

Time series is not stationary since there is a seasonal trend????
There is definitely a seasonal trend which repeats itself every 12 data points. This time series is a monthly time series. 


We first get rid of the seasonal trend by differencing the values at the same month of each year. Plot the seasonally differenced time series in the code block below. 
There is a clear upward trend. 
Clear upward trend violates stationarity.

4. ACF: Closers spikes: MA order
5. ACF: spikes around seasonal lags: SMA order
6. PACF: Closers spikes: AR order
7. PACF: spikes around seasonal lags: SAR order

```{r}
#acf2(diff(data))
#acf2(diff(log(data)))
acf2(diff(diff(log(data)),12),max.lag=30)
```



We de-trend the seasonally differenced time series by taking non-seasonal differencing, diff()

The significant partial autocorrelation coefficient at lag 12 suggests the order of seasonal AR term, $P\leq 1$.

Significant adjacent lags in PACF suggest the order of AR terms, $p\leq 2$. 

Significant adjacent lags in ACF suggest the order of MA terms, $q\leq 1$.

The significant autocorrelation coefficient at lag 12 suggests the order of seasonal MA term, $Q\leq 1$.


8. Fit few different models
9. Compare AIC, SSE, choose model with minimum AIC
10. Parsimony principle (order sum <= 6)

```{r}
d=1
DD=1
per=12
for(p in 1:2){
  for(q in 1:2){
    for(i in 1:2){
      for(j in 1:4){
        if(p+d+q+i+DD+j<=10){
          model<-arima(x=data, order = c((p-1),d,(q-1)), seasonal = list(order=c((i-1),DD,(j-1)), period=per))
          pval<-Box.test(model$residuals, lag=log(length(model$residuals)))
          sse<-sum(model$residuals^2)
          cat(p-1,d,q-1,i-1,DD,j-1,per, 'AIC=', model$aic, ' SSE=',sse,' p-VALUE=', pval$p.value,'\n')
        }
      }
    }
  }
}
```




11. Time plot, ACF, PACF, Ljung-Box test of residuals

```{r}
m<-sarima(data,0,1,1,0,1,1,12)
m
```

There is a systematic departure from linearity in QQ-plot which implies that residuals have a heavier tail compared to the Gaussian distribution. 


p-values from Ljung-Box test are high meaning that there is no significant autocorrelation left in the residuals. 


ACF shows no significant autocorrelation in the residuals. 


There is no strong evidence against the whiteness of the residuals. 


```{r}
m$ttable
```


```{r}
beta1 = m$ttable[1] 
beta2 = m$ttable[2]

m$fit$sigma2

beta1
beta2
beta1*beta2


pvalue1 = m$ttable[7]
pvalue2 = m$ttable[8]
pvalue1
pvalue2


```

p-values are 0.0008 and 0.0028 for MA and seasonal MA coefficients, respectively. 
The fact that they are both less than any reasonable significant level, both coefficients (terms) are significant. 



$$
SARIMA(0,1,1,0,1,1)_{12} \\
(1-B)(1-B^{12})X_t = (1+\beta_1 B)(1+\beta_{12}B^{12})Z_t \\
(1-B^{12}-B+B^{13})X_t = (1+\beta_{12}B^{12} + \beta_1 B + \beta_1 \beta_{12}B^{13} )Z_t \\
X_t = (B+B^{12}-B^{13})X_t + (1  + \beta_1 B  +\beta_{12}B^{12} + \beta_1 \beta_{12}B^{13} )Z_t \\
X_t = X_{t-1} + X_{t-12} - X_{t-13} + Z_t + \beta_1 Z_{t-1} + \beta_{12} Z_{t-12}  +  \beta_1 \beta_{12} Z_{t-13} \\

$$



```{r}
model<- arima(x=data, order = c(0,1,1), seasonal = list(order=c(0,1,1), period=12))
summary(model)
```

Forecast didn't work with sarima, only with arima model
```{r}
forecast(model)
```

```{r}
plot(forecast(model))
```

```{r}
a<-sarima.for(data,12,0,1,1,0,1,1,12)
a
plot.ts(c(data,a$pred), main='Monthly + Forecast', ylab='', col='blue', lwd=3)
```

point forecast for the number of accidental deaths in the March of 1979: 8315


